---
title: "project 2 assigment"
author: "Diana Garcia"
date: '2022-08-02'
output: html_document
---

## Caching the Inverse of a Matrix

Matrix inversion is usually a costly computation and their may be some benefit to caching the inverse of a matrix rather than compute it repeatedly. The assignment is to write a pair of functions that cache the inverse of a matrix. For that, Write the following functions:

### MakeCacheMatrix
This function creates a special "matrix" object that can cache its inverse.

```{r, echo=TRUE}

# Argument parameter: Matrix, Output: List
makeCacheMatrix <- function(x = matrix()){
    # Initialize the inverse value
    m <- NULL
    
    # Method to set the matrix
    set <- function(y){
        x <<- y
        m <<- NULL
    }
    
    # Method to get the matrix
    get <- function() x
    
    # Method to set the inverse of the matrix
    setinverse <- function(inverse) m <<- inverse
    
    # Method to get the inverse of the matrix
    getinverse <- function() m
    
    # Output list
    list(set = set, get = get,
         setinverse = setinverse,
         getinverse = getinverse)
}

```

### CacheSolve

This function computes the inverse of the special "matrix" returned by makeCacheMatrix above. If the inverse has already been calculated (and the matrix has not changed), then the cachesolve should retrieve the inverse from the cache.

```{r, echo=TRUE}

# Argument parameter: Matrix, Output: Inverse Matrix
cacheSolve <- function(x, ...){
    # Initialize a matrix that is the inverse of x matrix
    m <- x$getinverse()
    
    # Return a matrix if it is the inverse of x matrix
    if(!is.null(m)) {
        message("getting cached data")
        return(m)
    }
    
    # Get the matrix from the object
    data <- x$get()
    
    # Method to solve the inverse using matrix multiplication
    m <- solve(data, ...)
    
    # Set the inverse of inverse matrix
    x$setinverse(m)
    
    # Return the matrix
    m
}
```
